---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

Load and analyze activity data, per assignment.

First, some basic setup:

- Load required libraries
- Set echo=TRUE (per assignment)

```{r setup}
library(knitr)
library(plyr)
library(dplyr)
opts_chunk$set(echo=TRUE)
```

## Loading and preprocessing the data
Read data from activity.csv into **steps**, extracting from ZIP file if necessary:

```{r extract}
files <- list.files()
if (! ("activity.csv" %in% files)) {
    unzip("activity.zip")
}

steps <- read.csv("activity.csv", as.is=TRUE)
steps$interval <- as.factor(sprintf("%04d", steps$interval))
```

We also cast the interval as a factor that looks like "nnnn".

## What is mean total number of steps taken per day?

Aggregate the steps by day:

```{r stepsPerDay}
steps_by_day <- aggregate(x = steps[c("steps")],
                          FUN = "sum",
                          by = list(Group.date=steps$date))
```

Histogram of same:

```{r dailyStepsHistogram}
hist(steps_by_day$steps,
     breaks = 20,
     main = "Frequency of Steps per Day",
     xlab = "Steps in a Day")
```

Calculate mean and median steps per day:

```{r meanSteps}
mean_steps   <- mean(steps_by_day$steps, na.rm=TRUE)
median_steps <- median(steps_by_day$steps, na.rm=TRUE)
```

The mean number of steps per day is `r sprintf("%0.1f", mean_steps)`, and the median is `r median_steps`.

## What is the average daily activity pattern?

Break down steps by interval and plot:

```{r stepsByInterval}
steps_by_interval <- aggregate(steps ~ interval,
                               steps,
                               function(x) mean(x, na.rm = TRUE))

plot(steps_by_interval$interval,
     steps_by_interval$steps,
     type = "l",
     main = "Mean Steps through the Day",
     xlab = "Five-Minute Interval",
     ylab = "Steps")

# Lines don't seem to come out, so add this as backup.
lines(steps_by_interval$interval,
      steps_by_interval$steps)
```

Find the interval with the maximum number of steps:

```{r findMaxStepInterval}
max_steps <- max(steps_by_interval$steps)

max_interval <- steps_by_interval$interval[
    which.max(steps_by_interval$steps == max_steps)]
```

The interval with the highest number of steps, `r max_steps`,
begins at `r max_interval`.


## Imputing missing values

Count missing values:

```{r countMissingValues}
missing_value_count = sum(is.na(steps$steps))
```

There are `r missing_value_count` missing values in the original data set.

Fill in missing values with mean values for the five-minute interval:

```{r fillInMissingValues}
corrected <- steps # New data set

# Fill in missing values in new data set
for (i in 1:length(corrected$steps)) {
    if (is.na(corrected[i, "steps"])) {
        interval <- corrected[i, "interval"]
        interval_index <- which(steps_by_interval$interval == interval)
        fillin <- steps_by_interval[interval_index, "steps"]
        corrected[i, "steps"] <- fillin
    }
}
```

Calculate statistics on, and plot histogram of, "corrected" data set:

```{r showCorrectedValues}
corrected_steps_by_day <- aggregate(x = corrected[c("steps")],
                                    FUN = "sum",
                                    by = list(Group.date=corrected$date))

hist(corrected_steps_by_day$steps,
     breaks = 20,
     main = "Frequency of Steps per Day (Corrected)",
     xlab = "Steps in a Day")

# Calculate corrected mean and median.
corrected_mean_steps   <- mean(corrected_steps_by_day$steps, na.rm=TRUE)
corrected_median_steps <- median(corrected_steps_by_day$steps, na.rm=TRUE)

# Calculate percent change in corrected values from original values, for display.
mean_change_pct   <- (corrected_mean_steps / mean_steps - 1) * 100
median_change_pct <- (corrected_median_steps / median_steps - 1) * 100
```

The "corrected" mean number of steps is `r sprintf("%0.1f", corrected_mean_steps)`,
a `r sprintf("%0.3f", mean_change_pct)` change from `r sprintf("%0.1f", mean_steps)`.

The "corrected" median number of steps is `r sprintf("%0.1f", corrected_median_steps)`,
a `r sprintf("%0.3f", median_change_pct)` change from `r sprintf("%0.1f", median_steps)`.

## Are there differences in activity patterns between weekdays and weekends?

Add weekday/weekend indicator to data set:

```{r addTypeOfDayFactor}
# If day starts with "S", it's a weekend; otherwise a weekday.
steps$day_type <- factor(
    ifelse(
        grepl("^S", weekdays(as.POSIXlt(steps$date, "%Y-%m-%d")))),
        "weekend",
        "weekday")

steps_by_interval_and_day_type <- aggregate(steps ~ interval + day_type,
                                            steps,
                                            function(x) sum(x, na.rm = TRUE))

```